---
title: "Lab 3: Mammal Sleep Patterns"
author: "Merhan Ghandehari"
date: "February 6, 2016"
output: html_document
---



 ```{r, echo=FALSE}
# sales = read.csv("Data/Milwaukee_Sales_2012.csv")  # Load some data
# cor.test(sales$SalePrice, sales$FinSqft)  # Correlation test
# m = lm(SalePrice ~ FinSqft, data=sales)  # Fit a model
# plot(SalePrice ~ FinSqft, data=sales)  # Scatterplot...
# abline(m, col="red")  # ...with best fit line
```

```{r, echo=FALSE}
mammals = read.table("Data/SleepMammals.txt", header=TRUE)
summary(mammals)

# Converting the values for Predation, Exposure, and Danger from
# n umeric (integers) to categorical data
mammals$Predation = as.factor(mammals$Predation)
mammals$Exposure = as.factor(mammals$Exposure)
mammals$Danger = as.factor(mammals$Danger)
# Show a summary of these newly converted variables 
summary(mammals[, c("Predation", "Exposure", "Danger")])
```

	
Variable		  Description
BodyWt		    body weight (kg)
BrainWt		    brain weight (g)
NonDreaming		slow wave ("nondreaming") sleep (hrs/day)
Dreaming		  paradoxical ("dreaming") sleep (hrs/day)
TotalSleep		total sleep, sum of slow wave and paradoxical sleep (hrs/day)
LifeSpan		  maximum life span (years)
Gestation		  gestation time (days)
Predation		  predation index (1-5)
              1 = minimum (least likely to be preyed upon); 5 = maximum (most likely               to be preyed upon)
Exposure	  	sleep exposure index (1-5)
              1 = least exposed (e.g. animal sleeps in a well-protected den); 5 =                  most exposed
Danger		    overall danger index (1-5) (based on the above two indices and other                 information)
              1 = least danger (from other animals); 5 = most danger (from other                   animals)

```{r }
# cerate a new variable in mammals dataframe named BrainPct (the percentage of total body weight that the brain weighs)
# BrainPct = BrainWt / (BodyWt * 1000)
mammals$BrainPct = (mammals$BrainWt / (mammals$BodyWt * 1000))*100

#the highest BrainPct
HighestBrainPct = mammals[mammals$BrainPct == max(mammals$BrainPct), "Species"]
print(HighestBrainPct)

# HighestBrainPct = mammals[which.max(mammals$BrainPct), "Species"]


# the lowest BrainPct
LowestBrainPct = mammals[mammals$BrainPct == min(mammals$BrainPct), "Species"]
print(LowestBrainPct)

# the mean of BrainPct
MeanBrainPct = mean(mammals$BrainPct)
print(MeanBrainPct)

# the median of BrainPct
MedianBrainPct = median(mammals$BrainPct)
print(MedianBrainPct)
```


```{r, echo=FALSE, fig.align='center'}
# distribution of BrainPct
par(mfrow=c(2,2))
hist(mammals$BrainPct, col="gray", main = "Histgram of brain percentage", xlab = "Brain percentage")
plot(density(mammals$BrainPct), col="gray", main = "Density plot of brain percentage", xlab = "Brain percentage")
boxplot(mammals$BrainPct, col="gray", main = "Boxplot of brain percentage", ylab = "Brain percentage")

# vioplot is a combination of a density plot and a box plot
library(vioplot)
vioplot(mammals$BrainPct, col="gray")
title (main = "Vioplot of brain percentage", ylab = "Brain percentage")
```

- the distribution appears positively (right) skewed

```{r }
library(moments)
skewness(mammals$BrainPct)

library(e1071)
skewness(mammals$BrainPct)
```

The skewness here is 1.315525 or 1.283827. This value implies that the distribution of the data isskewed to the right or positively skewed.

3. a scatterplot with LifeSpan on the y-axis and Gestation on the x-axis

```{r }
plot(mammals$Gestation, mammals$LifeSpan, 
     xlab = "Gestation",
     ylab = "LifeSpan",
     pch = 20,
     col = "grey")
```
# get rid of all of the plots
# dev.off(dev.list()["RStudioGD"])

4. a boxplot that compares body weight across the five Danger levels.

```{r, echo=FALSE, fig.align='center'}
boxplot(mammals$BodyWt ~ mammals$Danger,
        col = "beige",
        xlab = "Danger Level = 1",
        ylab = "Body Weight", outline =F)


means = aggregate(BodyWt~Danger,data=mammals,FUN="mean", outline =F)
par(new=T)
points(means, col="red",pch=18)
```

4.2. a boxplot that compares TotalSleep across the five Danger levels

```{r, echo=FALSE, fig.align='center'}

boxplot(mammals$TotalSleep ~ mammals$Danger,
        col = "beige",
        xlab = "Danger Level",
        ylab = "Total Sleep",
        outline=FALSE)

means = aggregate(TotalSleep~Danger,data=mammals,FUN="mean", outline =F)
par(new=T)
points(means, col="red",pch=18)
```
- What inferences can you make about the sleeping conditions of an animal from this box plot?

???what is danger index?

# Data Transformations
5-
```{r }
mammals$LogBrainPct = log(mammals$BrainPct)
hist(mammals$LogBrainPct, freq=FALSE)
mu = mean(mammals$LogBrainPct)
std = sd(mammals$LogBrainPct)
curve(dnorm(x, mean=mu, sd=std), add=TRUE)  # Add curve to plot
```

6-
Data transformation is used the data does not have a normal distribution.

    There are many reasons for transformation. The list here is not
    comprehensive
        1. Convenience
        2. Reducing skewness
        3. Equal spreads
        4. Linear relationships
        5. Additive relationships
        
linear tranformation is used to change data from one sclae to another (such as z score), but it does not change the shape of the distribution
with transformation you can chnage an exponentiol corelation to a linear corelation (y=m^x -> log y = x log m -> log y = ax + b) so if we log y and see that we have a straight line we can conclude that the original data has been exponential. if it was not an staright line, you should log both x and y, and then if the scatter plot was straght then the original data is power 
        
# Correlation Analysis
7) correlation between lifespan and gestation period
```{r }
cor(mammals$LifeSpan, mammals$Gestation, use = "complete.obs", method = "pearson")

cor(mammals$LifeSpan, mammals$Gestation, use = "complete.obs", method =  "kendall")

cor(mammals$LifeSpan, mammals$Gestation, use = "complete.obs", method = "spearman")
```

Yes, from plot in question 3, it was clear that there is a positive linear relationship between these two variables.

8-
```{r }
MammalsSubset = data.frame(TotalSleep = mammals$TotalSleep, BodyWt = mammals$BodyWt, BrainWt = mammals$BrainWt, LifeSpan = mammals$LifeSpan, Gestation = mammals$Gestation)

# mammals[,sapply(mammals, is.numeric)] # this code is extract the numeric attributes, but here we don't need all of the numeric attributes

MammalsSubset_corr = cor(MammalsSubset, use = "complete.obs", method = "pearson") # get correlations

library('corrplot') #package corrplot
corrplot(MammalsSubset_corr, method = "circle", addCoef.col = "black") #plot matrix

# Use "Pairs Plot" from "psych" package to visualize scatter plot
library("psych")
pairs.panels(MammalsSubset, gap = 0)
```
Based on Pearson’s correlation coefficient, which two variables are most and least correlated with each-other?

Body weight and brain weight are the most correlated (0.93) and body weight and life span are the least correlated (0.30).

#Part II
## Linear Regression
9-
```{r }
# what shoud I do with NA; there are many NA in mammals$TotalSleep
Sleep_BodyWt.lm = lm(mammals$TotalSleep ~ mammals$BodyWt)
coeffs = coefficients(Sleep_BodyWt.lm); coeffs

#How to write the formula????
β0 = 10.835131637
β1 = -0.001524093

# coefficient of determination of a linear regression model (the variances of the fitted values and observed values of the dependent variable)  "Values of R2 close to 1 imply that most of the variability in y is explained by the regression model." "0% indicates that the model explains none of the variability of the response data around its mean." "Metric for evaluating the goodness of fit of your model. Higher is better with 1 being the best. "

"*** for high significance and * for low significance"
summary(Sleep_BodyWt.lm)$r.squared 

summary(Sleep_BodyWt.lm)
# P-value is higher than .05. so we accept the null hypothesis; it means that there is not a significant relationship between the variables in the linear regression model "The p-value for each term tests the null hypothesis that the coefficient is equal to zero (no effect). A low p-value (< 0.05) indicates that you can reject the null hypothesis. In other words, a predictor that has a low p-value is likely to be a meaningful addition to your model because changes in the predictor's value are related to changes in the response variable. Conversely, a larger (insignificant) p-value suggests that changes in the predictor are not associated with changes in the response."


plot(mammals$BodyWt, mammals$TotalSleep,
     main = "XXXX",
     xlab = "XXX",
     ylab = "YYY",
     pch = 20,
     col = "grey")
# Add fit lines
# Linear regression line (y ~ x) 
abline(Sleep_BodyWt.lm, col="red")

par(mfrow=c(2,2))
plot (Sleep_BodyWt.lm)

# Residual Plot
Sleep_BodyWt.res = resid(Sleep_BodyWt.lm)
plot(mammals$BodyWt[1:58], as.vector(Sleep_BodyWt.res), 
   ylab="Residuals", 
   xlab="BodyWt", 
   main="XXX") 
abline(0, 0) # the horizon
```
10- similar to 9, just by replacing variables
```{r }
Sleep_BodyWt.lm = lm(mammals$TotalSleep ~ mammals$BrainWt)
coeffs = coefficients(Sleep_BodyWt.lm); coeffs
summary(Sleep_BodyWt.lm)

plot(mammals$BrainWt, mammals$TotalSleep,
     main = "XXXX",
     xlab = "XXX",
     ylab = "YYY",
     pch = 20,
     col = "grey")
# Add fit lines
# Linear regression line (y ~ x) 
abline(Sleep_BodyWt.lm, col="red")
```
##Multiple Regression
```{r }
MultiReg.lm = lm(mammals$TotalSleep ~ mammals$BodyWt + mammals$BrainWt + mammals$LifeSpan + mammals$Gestation)

summary(MultiReg.lm)
```




```{r }
# what shoud I do with NA; there are many NA in mammals$TotalSleep
Sleep_BodyWt.lm = lm(mammals$TotalSleep ~ mammals$BodyWt)
coeffs = coefficients(Sleep_BodyWt.lm); coeffs

test = na.omit(data.frame(TotalSleep=mammals$TotalSleep, BodyWt=mammals$BodyWt))
test2 = test[test$TotalSleep > quantile(test$TotalSleep, .25) - 1.5*IQR(test$TotalSleep) & test$TotalSleep < quantile(test$TotalSleep, .75) + 1.5*IQR(test$TotalSleep) & test$BodyWt > quantile(test$BodyWt, .25) - 1.5*IQR(test$BodyWt) & test$BodyWt < quantile(test$BodyWt, .75) + 1.5*IQR(test$BodyWt),]

Sleep_BodyWt.lm = lm(test2$TotalSleep ~ test2$BodyWt)
summary(Sleep_BodyWt.lm)

plot(test2$BodyWt, test2$TotalSleep,
     main = "XXXX",
     xlab = "XXX",
     ylab = "YYY",
     pch = 20,
     col = "grey")
# Add fit lines
# Linear regression line (y ~ x) 
abline(Sleep_BodyWt.lm, col="red")
```
